---
title: "IFNbeta LPS IL6 CD4 counts microbiome family"
author: "Guannan Shen"
date: "April 23, 2019"
output: 
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
---

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
opts_chunk$set(engine = "R")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
## setting wd in DELL
## opts_knit$set(root.dir = "~/Stats/CIDA_OMICs/CIDA_OMICS/7659Stats_Genetics/HW5/")
## setting working directory in asus 
## opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/gitlab/Cario_RNASeq_Microbiom_Inte/DataRaw/") 
## setting working directory in ubuntu
opts_knit$set(root.dir = "~/Documents/gitlab/Omics_Integration/DataRaw/hiv_infected_un/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```



```{r libs, echo=FALSE }
## set up workspace
rm(list = ls())

library(knitr)
library(tidyverse)
library(SmCCNet)
library(mixOmics)
library(r.jive)
library(openxlsx)
library(stats)
library(readxl)

options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
## not in function
'%nin%' <- Negate('%in%')

# ######## clean memory ######################
# rm(list = ls())
# gc()
# ls()
# slotNames(x)
# getSlots(x)

```

# Omics Integration and Network Analysis

## Data
Clinical parameters: IL-6 (pg/ml), LPS (pg/ml) and CD4 count (cells/ul).  
Family level microbiome data, filtered via tidi_MIBI.  
RNAseq data of IFN-Beta genes list, TMM normalized.  
27 subjects in common.  
423 genes and 53 families. 

```{r importdata}
## clinical data 
## IL6 lps cd4_counts
clin_micro <- read.xlsx("clin_microbiome.xlsx")

## rna seq data 
beta <- read.table("genesbeta", header = T)
colnames(beta)

# test if the Symbols are unique
nrow(beta) == length(table(beta[,2]))
tmm <- read.xlsx("TMM_normalized_counts.xlsx")
tmm_beta <- data.frame(tmm) %>% dplyr::filter(Gene_ID %in% beta$Gene_ID ) %>% 
           dplyr::select(-Gene_ID) %>% 
                 column_to_rownames("Symbol") %>% t %>% as.data.frame

# change the ID
tmm_beta_id <- c(gsub("C", "MIHIV", row.names(tmm_beta)  [1:13]  ),
                gsub( "H", "MIHIV", row.names(tmm_beta)  [14:32] ) )
sum(tmm_beta_id %in% clin_micro$ID)
row.names( tmm_beta) <- tmm_beta_id

# family level microbiome data
fam_tidi <- read.csv('family_mibiset.csv')
fam_clr <- data.frame(fam_tidi) %>% dplyr::select(Lib, Taxa, clr) %>%
            spread(., Taxa, clr) %>% dplyr::mutate(ID = gsub( "B", "",  .$Lib)) %>%
             dplyr::select(-Lib) %>%  dplyr::select(ID, everything())
# list of common subject
sum(fam_clr$ID[fam_clr$ID %in% tmm_beta_id] %nin%
tmm_beta_id[tmm_beta_id %in%  fam_clr$ID])

# ids
IDs <- fam_clr$ID[fam_clr$ID %in% tmm_beta_id]
############## SmCCNet input #################
# subject at rows, feature at columns 
# filter by ID
# mean 0, unit variance 
# One way to normalize variables in R is to use the scale function
## clin
clin_27 <- data.frame(clin_micro) %>%  dplyr::filter(ID %in% IDs) 

il6_cd4_lps <- cbind(clin_27[, 1:4], data.frame(apply(clin_27[,c(5:6, 11)], 2, scale)) )

#### microbiome 
fam_sm <- fam_clr %>% dplyr::filter(ID %in% IDs) %>% column_to_rownames("ID")
fam_scale <-  data.frame(apply(fam_sm, 2, scale))
fam_scale$ID <- row.names(fam_sm)


## tmm

beta_tmm_scale <- data.frame(apply(tmm_beta[row.names(tmm_beta) %in% IDs ,], 2, scale))
beta_tmm_scale$ID <- row.names(tmm_beta)[row.names(tmm_beta) %in% IDs]

## merge dataset
data <- merge( fam_scale, 
merge( il6_cd4_lps,beta_tmm_scale, by = "ID" ),
by = "ID")

```

## Methods
1. SmCCNet.  
2. mixOmics.  
3. r.jive.  

```{r smccnet}
# fam_sm, 53 families
# clinical data data[, 55:60], gender, age, hiv, cd4, il6, lps
# tmm_beta, 423 genes 
dim(data)
microbiome <- data[, 2:54]
#  clin
# il6
il6 <- data[, 59]
#
gene <- data[, 61:483]
# n for p1 and all names
n <-  ncol(gene)
names <- c(colnames(gene), colnames(microbiome))
# corre
x <- cbind(data[, 61:483], data[, 2:54])
corr <- cor(x)

#######  Unweighted SmCCA
# SubsamplingNum Number of feature subsamples. Default is 1000. Larger number leads to more
# accurate results, but at a higher cost.
# this sample size, 27, the paper use 4-fold 
set.seed(123)
W1 <- getRobustPseudoWeights(gene, microbiome, Trait = il6, Lambda1 = 0.01,
      Lambda2 = 0.01, s1 = 0.7, s2 = 0.9, NoTrait = FALSE, FilterByTrait = FALSE,
      SubsamplingNum = 2000, CCcoef = NULL, trace = FALSE)
dim(W1)
abar <- getAbar(W1, P1 = n, FeatureLabel = names)
modules <- getMultiOmicsModules(abar, P1 = n, CutHeight = 1 - 0.1^10, PlotTree = TRUE)
plotMultiOmicsNetwork(abar, corr, modules, ModuleIdx = 1, P1 = n, FeatureLabel = names)
## other
cd4 <- data[, 58]
lps <- data[, 60]
# c("cd4", "lps"), as.name(), eval()

W1 <- getRobustPseudoWeights(gene, microbiome, Trait = cd4, Lambda1 = 0.01,
      Lambda2 = 0.01, s1 = 0.7, s2 = 0.9, NoTrait = FALSE, FilterByTrait = FALSE,
      SubsamplingNum = 2000, CCcoef = NULL, trace = FALSE)
dim(W1)
abar <- getAbar(W1, P1 = n, FeatureLabel = names)
modules <- getMultiOmicsModules(abar, P1 = n, CutHeight = 1 - 0.1^10, PlotTree = TRUE)
plotMultiOmicsNetwork(abar, corr, modules, ModuleIdx = 1, P1 = n, FeatureLabel = names)

W1 <- getRobustPseudoWeights(gene, microbiome, Trait = lps, Lambda1 = 0.01,
      Lambda2 = 0.01, s1 = 0.7, s2 = 0.9, NoTrait = FALSE, FilterByTrait = FALSE,
      SubsamplingNum = 2000, CCcoef = NULL, trace = FALSE)
dim(W1)
abar <- getAbar(W1, P1 = n, FeatureLabel = names)
modules <- getMultiOmicsModules(abar, P1 = n, CutHeight = 1 - 0.1^10, PlotTree = TRUE)
plotMultiOmicsNetwork(abar, corr, modules, ModuleIdx = 1, P1 = n, FeatureLabel = names)

```


## Questions
1. Weighted SmCCA might be better, need to define the weights.  
2. Lasso penalty, CV to decide the Lambdas.  
3. All the inputs, TMM normalized counts, centered log ratio transform of microbiome relative abundance and phenos were scaled to mean 0, unit variance.   
4. Set the random seed, should the results be robust to random seed?  
5. The canonical correlation weight matrix are mostly zeros.  