---
title: "Cario phylum family genus species"
author: "Guannan Shen"
date: "May 7, 2019"
output: 
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: no
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
---

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
# opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
# opts_chunk$set(engine = "R")
# knitr::opts_chunk$set(echo = T)
# knitr::opts_chunk$set(message = F)
# knitr::opts_chunk$set(warning = F)
## setting wd in DELL
## opts_knit$set(root.dir = "~/Stats/CIDA_OMICs/CIDA_OMICS/7659Stats_Genetics/HW5/")
## setting working directory in asus 
## opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/gitlab/tidi_MIBI/Data/") 
## setting working directory in ubuntu
opts_knit$set(root.dir = "~/Documents/gitlab/tidi_MIBI/Data/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```



```{r libs, echo = FALSE, warning= F}
######################
## Set up workspace
######################
rm(list = ls())
library(openxlsx)
library(tidyverse)


options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
## not in function
'%nin%' <- Negate('%in%')

# ######## clean memory ######################
# rm(list = ls())
# gc()
# is(dds)
# slotNames(dds)
```


# Data
Focus on the Phylum, Family and Genus level. 

Because, in our case, the OTUs picking was done by reference database, thus we don’t have complete taxanomy at the species level. As a result, for alpha diversities calculation, we can only go down to the genus level.

## Negative Control: MIHIVH20 with 80K counts
The negative control is not in OTU tables data files. 

```{r importdata}
############ set the folder ########
# asus
# folder <- c('C:/Users/hithr/Documents/Stats/gitlab/tidi_MIBI/')
# source('C:/Users/hithr/Documents/Stats/gitlab/tidi_MIBI/Code_charlie/tidi_MIBI.R')
# ubuntu
folder <- c("~/Documents/gitlab/tidi_MIBI/")


## source 
source( paste(folder, "Code_charlie/tidi_MIBI.R", sep = "") )

## 
levels <- data.frame(matrix( data = c("Phylum", "Class", "Order", "Family", "Genus", "Species") , ncol = 1))
colnames(levels) <- "Levels of Microbiome Data"
kable(levels)

#
read_data <- function(folder, datafile, header){
  read.table(paste(folder, datafile, sep = ""), header = header)
}

############## filter the data #############
#### filter the sample by the shared set of samples
# the naming convention, a B at the end of the name
filter_micro <- function(data, samples){
  data[1,-1] = gsub("B", "", data[1,-1])
  name = c("OTU_Name", samples)
  return(data[, (data[1,] %in% name)]  )
}

###### having column names#######
filter_micro_t <- function(data, samples){
  colnames(data)[-1] = gsub("B", "", colnames(data)[-1])
  name = c("OTU_Name", samples)
  return(data[,  colnames(data) %in% name]  )
}

## shared sample size across datasets
shared_sam <- c( "MIHIV124", "MIHIV132", "MIHIV138", "MIHIV154", "MIHIV178", "MIHIV255", "MIHIV278", "MIHIV286", "MIHIV323", "MIHIV361", "MIHIV391", "MIHIV404", "MIHIV428", "MIHIV493", "MIHIV582", "MIHIV594", "MIHIV648", "MIHIV683", "MIHIV708", "MIHIV716", "MIHIV819", "MIHIV825", "MIHIV839", "MIHIV914", "MIHIV947", "MIHIV972", "MIHIV998")

# fam %>% filter_micro(., shared_sam)

#################### The full set of data ##############
phy_all <- read_data(folder, "Data/phylum_biopsy_cts_24Aug2018.txt", F)
# ord <- read_data(folder, "Data/order_biopsy_cts_24Aug2018.txt") 
# fam <- read_data(folder, "Data/family_biopsy_cts_24Aug2018.txt")
# gen <- read_data(folder, "Data/alltaxa_biopsy_cts_24Aug2018.txt")
# spe <- read_data(folder, "Data/species_biopsy_cts_24Aug2018.txt") 

############ The shared sample across datasets #############
phy <- read_data(folder, "Data/phylum_biopsy_cts_24Aug2018.txt", F) %>% filter_micro(., shared_sam)
ord <- read_data(folder, "Data/order_biopsy_cts_24Aug2018.txt", F) %>% filter_micro(., shared_sam)
fam <- read_data(folder, "Data/family_biopsy_cts_24Aug2018.txt", F) %>% filter_micro(., shared_sam)
gen <- read_data(folder, "Data/alltaxa_biopsy_cts_24Aug2018.txt", F) %>% filter_micro(., shared_sam)
spe <- read_data(folder, "Data/species_biopsy_cts_24Aug2018.txt", F) %>% filter_micro(., shared_sam)

# 
# ## Meta Data to merge with OTU table
# old.meta <- read.table("C:/Users/hithr/Documents/Stats/gitlab/tidi_MIBI/Data/SEQ020_Wilson1_metadata_all_subjects_24Aug2018.txt", header = T ) 

old.meta <- read.table(paste(folder, "Data/SEQ020_Wilson1_metadata_all_subjects_24Aug2018.txt", 
                             sep = ""), 
                       header = T ) 


## colnames(old.meta)
paste("Samples in Microbiome Data")
sample <- gsub( "B", "", old.meta$Lib)
sample
"MIHIVH20" %in% sample

sum(gsub( "B", "", phy_all[1,-1]) %nin% sample)
paste("The OTU data and Clinical Data have the same set of samples.")

########### shared samples ##################
paste("Samples shared across Datasets")
meta <- old.meta %>% dplyr::mutate(Lib = sample) %>% dplyr::filter(Lib %in% shared_sam)
paste("Sample Size:", length(table(meta$Lib)))



```

# Exploratory Data Analysis

## Library Size 

```{r library_size}
 ########## merge the data files ############
# phylum data with column names 
library_size <- read_data(folder, "Data/phylum_biopsy_cts_24Aug2018.txt", T)  %>% 
           filter_micro_t(., shared_sam) %>% .[1,-1] %>% t %>% as.data.frame
library_size$Lib <- row.names(library_size)
meta_libsize <- merge(meta, library_size, by = "Lib")
##### check alignment 
sum(meta_libsize$Lib != phy[1,-1])

## 
lib_size <- function(data){
  x1 = as.numeric(as.vector(unlist(data[2,-1])))
  name = gsub( "B", "", data[1,-1])
 
  p1 = ggplot(mapping = aes(x1 )) +
      geom_histogram (stat = "bin", bins = 50) +
      labs(caption = paste("Library Size"), x = "Library Size" ) +
      theme_bw()
  print(p1)
  p2 = ggplot(mapping = aes(name, x1,  fill = meta_libsize$diagnosis )) +
      geom_bar (stat="identity") +
      labs(caption = paste("Library Size"), x = "Samples" , y = "Total Counts") +
      coord_flip() +
      ## theme(axis.text.x = element_text(angle = 90)) + 
      theme_bw()
  print(p2)
  p3 = ggplot(mapping = aes(name, x1, fill = meta_libsize$Sex )) +
      geom_bar (stat="identity") +
      labs(caption = paste("Library Size"), x = "Samples" , y = "Total Counts") +
      coord_flip() +
      ## theme(axis.text.x = element_text(angle = 90)) + 
      theme_bw()
  print(p3)
  p4 = ggplot(mapping = aes(name, x1, fill = meta_libsize$Race )) +
      geom_bar (stat="identity") +
      labs(caption = paste("Library Size"), x = "Samples" , y = "Total Counts") +
      coord_flip() +
      ## theme(axis.text.x = element_text(angle = 90)) + 
      theme_bw()
  print(p4)
}

lib_size(phy)

##
meta_libsize %>% dplyr::select(Lib, `1`, diagnosis, Sex, Age, Race, sex_dx) %>% 
        arrange(`1`) %>% kable(col.names = c("ID", "Library Size", "Diagnosis", "Sex",
                                             "Age", "Race", "Sex_dx"))



```




## Covariates
Predictors usually are included in the model are, "diagnosis", "Sex", "Age", "Race", "sex_dx". 

```{r covariates}
## 
ggplot(meta, aes(diagnosis, fill = diagnosis)) +
  geom_bar() +
  theme_bw()

ggplot(meta, aes(Race, fill = Race)) +
  geom_bar() +
  theme_bw()

ggplot(meta, aes(Sex, fill = Sex)) +
  geom_bar() +
  theme_bw()

ggplot(meta, aes(sex_dx, fill = sex_dx)) +
  geom_bar() +
  theme_bw()

ggplot(meta, aes(Age)) +
  geom_histogram(bins = 20) +
  theme_bw()
```

## Relative Abundance (RA) and Prevalence
At phylum, family, genus and species levels. 

Taxa which are present in a minority of the population were collapsed into a single “other” category. 

```{r RA, fig.height=12, fig.width=10}
phy_t <- read_data(folder, "Data/phylum_biopsy_cts_24Aug2018.txt", T)  %>% 
           filter_micro_t(., shared_sam)
ord_t <- read_data(folder, "Data/order_biopsy_cts_24Aug2018.txt", T) %>% 
  filter_micro_t(., shared_sam)
fam_t <- read_data(folder, "Data/family_biopsy_cts_24Aug2018.txt", T) %>% 
  filter_micro_t(., shared_sam)
gen_t <- read_data(folder, "Data/alltaxa_biopsy_cts_24Aug2018.txt", T) %>% 
  filter_micro_t(., shared_sam)
spe_t <- read_data(folder, "Data/species_biopsy_cts_24Aug2018.txt", T) %>% 
  filter_micro_t(., shared_sam)

get_RA <- function(data){
 ######## pull out information from data #############
  lib_s = data[1,-1]; samples = colnames(data)[-1]
  taxa = data[-1, 1]; data_core = data[-1,-1]
  #### prevalence ####
 non_zero = apply(data_core , 1, function(x){
  sum(x != 0)
   } ) %>% as.data.frame() %>% .[,1]

  ####### calculate RA ############
 data_ra = matrix(NA, nrow = nrow(data_core))
 for( i in 1:ncol(data_core) ){
  ra = round(data_core[,i]/ as.numeric(lib_s[i] ) *100, 4)
  data_ra = cbind(data_ra,  ra)
 }
  data_ra = data.frame( data_ra[, -1] )
 row.names( data_ra ) = taxa
  colnames( data_ra ) = samples
 #### avg_RA #####
  avg_ra = apply(data_ra, 1, mean) %>% as.data.frame() %>% .[,1] %>% round(., 4)

  res = data_ra  %>% dplyr::mutate(Taxas = taxa,
                          Prevalence = non_zero,
                           Avg_RA = avg_ra)  %>% 
            arrange( desc(Prevalence), desc(Avg_RA)) %>% 
             dplyr::mutate(Num = 1:nrow(.)) %>% 
               dplyr::select(Num, Taxas, Prevalence, Avg_RA, everything())
  return(res)
}

phy_ra = get_RA(phy_t)
phy_ra %>% kable(caption = "Phyla")

ord_ra = get_RA(ord_t)
ord_ra %>% kable(caption = "Orders")

fam_ra = get_RA(fam_t)
fam_ra %>% kable(caption = "Families")

gen_ra = get_RA(gen_t)
gen_ra %>% kable(caption = "Genera")

spe_ra = get_RA(spe_t)


############# box plot of relative abundance ###########


####### same order as meta_libsize ########### 
sum(row.names(
phy_ra %>% dplyr::select(-Num, -Prevalence, -Avg_RA) %>% 
     column_to_rownames("Taxas") %>% t() %>% as.data.frame()
) != meta_libsize$Lib )

stack_boxplot <- function(data_ra, xlab){
  data = data_ra %>% dplyr::select(-Num, -Prevalence, -Avg_RA) %>% 
    ##################### the stack function ###################
     column_to_rownames("Taxas") %>% t() %>% as.data.frame() %>% stack()
  ggplot(data, aes(x = ind, y = values)) +
  geom_boxplot() + 
  labs(x = xlab, y = "Relative Abundance") +
  theme(axis.text.x = element_text(angle = 90, size = 10)) +
  ## geom_dotplot(binaxis='y', stackdir='center', dotsize=.1) 
  geom_jitter(shape=16, position=position_jitter(0.2), size = 1)
}
  
stack_boxplot(phy_ra, "Phyla")
stack_boxplot(ord_ra, "Orders")

stack_boxplot_n <- function(data_ra, xlab, n){
  data = data_ra %>% dplyr::select(-Num, -Prevalence, -Avg_RA) %>% 
    ##################### the stack function ###################
     column_to_rownames("Taxas") %>% t() %>% as.data.frame() %>% .[, 1:n] %>% stack()
  name = paste(xlab, "Top", n)
  ggplot(data, aes(x = ind, y = values)) +
  geom_boxplot() + 
  labs(x = name, y = "Relative Abundance") +
  theme(axis.text.x = element_text(angle = 90, size = 10)) +
  ## geom_dotplot(binaxis='y', stackdir='center', dotsize=.1) 
  geom_jitter(shape=16, position=position_jitter(0.2), size = 2)
}

stack_boxplot_n(ord_ra, "Orders", 20)
stack_boxplot_n(fam_ra, "Families", 20)
stack_boxplot_n(gen_ra, "Genera", 20)
stack_boxplot_n(spe_ra, "Species", 20)


```



## Alpha Diversity Measures


```{r alpha, fig.height=12, fig.width=10}
ord_set0 <-  tidi_MIBI(class = ord, # OTU Table
                      meta = meta, ## Meta Data
                      prev_cutoff = 0, ## Prevalence cutoff
                      ra_cutoff = 0,   ## Relative abundance cutoff
                      unc = TRUE)  

mibi.set <-  tidi_MIBI(genus = gen, # OTU Table
                      meta = meta, ## Meta Data
                      ## Prevalence cutoff 0
                      prev_cutoff = 0, 
                      ## Relative abundance cutoff 0 
                      ra_cutoff = 0,   
                      unc = TRUE)  
gen_set0 <- mibi.set %>% 
          otu_filter(rank = "Genus", 
              prev_cutoff = 0, 
              ra_cutoff = 0,   
              unc = TRUE)  

########### Alpha Diversity

all_gen_alpha <- Alpha_Div(gen_set0, rank = "Genus")
## same order
sum(all_gen_alpha[1:27, 2] != meta_libsize$Lib)

data_alpha <- all_gen_alpha[1:27, 9:14]  %>% 
  as.data.frame() %>% stack() %>% dplyr::mutate(ID = rep(all_gen_alpha[1:27, 2], 6),
                                               Diagnosis =  rep(meta_libsize$diagnosis,6),
                                               Sex = rep(meta_libsize$Sex, 6),
                                               Race = rep(meta_libsize$Race, 6),
                                               sex_dx = rep(meta_libsize$sex_dx, 6))


all_gen_alpha[1:27, 2]
## 
ggplot(data_alpha, aes(x = ID, y = values)) +
  facet_wrap(~ind, scales = "free") +
  geom_point(size = 3, aes(colour = Diagnosis, shape = Sex)) +
  theme(axis.text.x = element_text( size = 3)) + 
  theme_bw() +
  labs(x = "Samples", y = "Alpha Diversity Measures")

ggplot(data_alpha, aes(x = ID, y = values)) +
  facet_wrap(~ind, scales = "free") +
  geom_point(size = 3, aes(colour = Race, shape = sex_dx)) +
  theme(axis.text.x = element_text( size = 3)) + 
  theme_bw() +
  labs(x = "Samples", y = "Alpha Diversity Measures")
 


```

Error: quantile(as.numeric(x), c(0.25, 0.75), na.rm = na.rm, names = FALSE, : object 'mibi.set' not found. 